---
title: "Assignment1"
author: "phseven"
date: "July 11, 2014"
output: html_document
---
Input: activity.csv file (assumed to be in same directory as this R script)

## Load and preprocess the data.

1. Load the data into a datatable ad.
2. Change the dt column class to IDate from character.
3. Set (dt, intv) as key to the datatable.

```{r, echo=TRUE}
require(data.table)
require(ggplot2)

# Read activity.csv into a datatable ad with fread. 

ad <- fread("activity.csv")

# Set data table column names.

setnames(ad, c("step", "dt", "intv"))

# Change dt class from character to IDate

ad[, dt := as.IDate(dt, format="%Y-%m-%d")]

# Set date and interval as key to the datatable.

setkey(ad, dt, intv)
```

## Calculate total number of steps taken per day.

1. Make a histogram of total number of steps per day.
2. Calculate mean and median total number of steps per day.

```{r, echo=TRUE}

# Sad is summary datatable containing total number of step (tot.step)
# grouped by dt.

sad <- ad[, list(tot.step = sum(step, na.rm=TRUE)), by=dt]

# Plot histogram.

qplot(dt, tot.step, data=sad, geom="histogram", main="Steps Per Day",
      stat="identity" , xlab="Date", ylab="Number of Steps")

# Calculate mean and median total number of steps per day.

mm1 <- sad[, list(Mean = mean(tot.step), Median = median(tot.step))]
```

## Total number of steps:
- Mean = `r mm1$Mean`
- Median = `r mm1$Median`

## Average daily activity pattern


Plot a time series of average number of steps taken in each interval.
First, create a datatable mad with 2 columns: intv and avg.step. Intv will be
the key and avg.step will contain no. of steps averaging over all days for
that particular interval.

```{r, echo=TRUE}

# mad datatable will have two columns: intv, avg.step; intv will be the key.

mad <- ad[, list(avg.step = mean(step, na.rm=TRUE)), keyby=intv]

# Create the x-axis tick vector

xax <- c( "12am", paste0(seq(2,10,2), "am" ), "12pm", 
          paste0(seq(2,10,2), "pm"), "12am ")

# Plot the average no. of steps at different hours of the day graph.
# Note intv (x-value) is scaled and shifted to correspond to the x-axis
# ticks.

qplot(1+intv/200, avg.step, data=mad, geom="line", 
      main="Average Number of Steps at Different Hours of the Day",
      stat="identity" , xlab="Time", 
      ylab="Average Number of Steps (averaged over 5 minute intervals)"
      ) + 
    xlim(xax) + 
    geom_vline(xintercept=(1 + 835/200), linetype="longdash", color="blue")


# Time interval with the maximum no. of steps (on average)

mad[avg.step == max(avg.step), list(Max.Avg.Step.Interval = intv, 
                                    Max.Avg.Step = avg.step)]

```

Make a copy of the ad (activity datatable) to fad.
Fill the NA values of fad$step with mad$avg.step for that interval. Note, 
however, fad$step is integer, whereas mad$avg.step is numeric, so mad$avg.step
will be converted to integer using the as.integer function.

```{r, echo=TRUE}

# Make a copy of ad datatable to fad.

fad <- copy(ad)

# Set intv as the key for fad, so that it can be joined with mad.

setkey(fad, intv)

# If fad$step is NA, update it with mad$avg.step, 
# else update with its old value.

fad[mad, step := ifelse(is.na(step), as.integer(avg.step), step)]

```

Plot a new histogram of total number of steps per day after imputing 
missing values. 

Calculate mean and median of total number of steps per day.

```{r, echo=TRUE}

sfad <- fad[, list(tot.step = sum(step)), by=dt]

qplot(dt, tot.step, data=sfad, geom="histogram", 
      main="Steps Per Day (after imputing missing data)",
      stat="identity" , xlab="Date", ylab="Number of Steps")
       
sfad[, list(Mean = mean(tot.step), Median = median(tot.step))]

```

Plot a graph of total number of steps per day conditioning on weekday vs
weekends. 

Create a new column wdy in fad datatable which will contain a factor
with two possible values - Weekday or Weekend.

Calculate mean and median of total number of steps per day.

```{r, echo=TRUE}

# Function wday() returns 1 = Sun, 2=Mon...7=Sat.
# Create new column wdy in fad datatable.

fad[, wdy := factor(ifelse(wday(dt) %in% c(1, 7), 0L, 1L),
                    levels= c(0L, 1L),
                    labels= c("Weekend", "Weekday"), ordered=TRUE)]

# Create a new summary table wsfad based on fad grouped by wdy and intv.

wsfad <- fad[, list(avg.step = mean(step)), by = list(intv, wdy)]

qplot(intv, avg.step, 
      data=wsfad, geom="line", 
      main="Average Number of Steps vs Interval",
      stat="identity", xlab="Interval", 
      ylab="Average Number of Steps (averaged over 5 minute intervals)"
      ) + facet_wrap(~ wdy, nrow=2) + geom_line(colour="blue")
       
sfad[, list(Mean = mean(tot.step), Median = median(tot.step))]

```





Function to convert interval to ITime. The unit of interval is 1/100 hr
(so for each day, there are intervals from 0..2355, and each interval spans
5 units).

```{r, echo=TRUE}
to.Time <- function(x) {
    hr <- floor(x / 100)
    min <- x %% 100
    hm(sprintf("%02d:%02d", hr, min))
}
```


Open a png device, plot 4 graphs and close the device.

```{r, echo=TRUE}
#png(file = "plot4.png", width=480 , height=480)

setkey(ad, DT, TM)

par(mfrow= c(2,2))

# plot1

ad[, plot(strptime(paste(DT, TM), format="%d/%m/%Y %T"), 
    AP, type="l", xlab="",
    ylab="Global Active Power (kilowatts)") ]

# plot2

ad[, plot(strptime(paste(DT, TM), format="%d/%m/%Y %T"), 
    Volt, type="l", xlab="datetime",
    ylab="Voltage") ]

# plot3

ad[, plot(strptime(paste(DT, TM), format="%d/%m/%Y %T"), 
    SM1, type="n", 
    xlab="", ylab="Energy sub metering") ]

legend("topright", 
       legend= c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"),
       lty="solid", lwd=2, bty="n",
       col= c("black", "red", "blue"))

ad[, lines(strptime(paste(DT, TM), format="%d/%m/%Y %T"), 
    SM1, type="l", col="black")]

ad[, lines(strptime(paste(DT, TM), format="%d/%m/%Y %T"), 
    SM2, type="l", col="red")]

ad[, lines(strptime(paste(DT, TM), format="%d/%m/%Y %T"), 
    SM3, type="l", col="blue")]

# plot4

axisTicks(c(0.0,0.5), log=FALSE, axp=c(0.0,0.5,5))
ad[, plot(strptime(paste(DT, TM), format="%d/%m/%Y %T"), 
    RP, type="l", xlab="datetime",
    ylab="Global_reactive_power") ]

#dev.off()
```
